plyranges: a grammar for transforming genomics data

### 2018-03-21 paper.Rmd
intro polish

There is a large variety of genomics data available to the modern biologist
from sequencing assays to public collections: annotations of genomic features
such as genes or exons for reference genomes, coverage vectors in the BigWig
format, peaks called from ChIP-Seq experiments and many more.
All of these hetergenous data types can be naturally represented as annotated sets of 
integer pairs, corresponding to the start and end points of sequences, along with 
strandedness and chromosome location to provide biological context. Measurements
or metadata can be included as additional features on their relevant sequences. 
As an example agene can be represented with its coordinates, along with its 
identifier and the identifiers of its exons; or an RNA-seq experiment may be
represented as sets of genes with a matching count column.

@@

The Bioconductor/R packages `IRanges` and `GenomicRanges` 
define a DSL for analysing genomics data with R via
an object-oriented (OO) interface @Lawrence2013-wg; @Huber2015-ei. This interface
provides abstractions for representing genomics data and enables interoperability
by allowing users and developers to use these abstractions in their own code
and packages. However, due to the large number of abstractions there are costs
involved. Due to the OO interface each abstraction will behave slightly 
differently for any given generic method. There is a cognitive load on the user
to understand the appropriate context for each abstraction. An alternative approach
is to recognise that most genomic data analysis tasks can be performed with 
data structures that naturally represent genomic intervals.

@@

The tidy data abstraction enables the construction of fluent application programming interfaces (APIs) [@Fowler2010-zd] that are optimised for readability. An example of a 
fluent interface built on the tidy data pattern is the R pacakge `dplyr` [@Wickham2017-dplyr]. 
The package defines a consistent syntax and style for transforming tabular data sets via
functions that are named as verbs. Each function performs an action on the data
set and is cohesive and endomorphic. Cohesion means that each function performs
a singular task and has no side-effects, while endomorphism means a function
call returns the same type of object as its input.

Then the last paragraph (already pretty close) states that:

    we aim to simplify the interface to the most fundamental Ranges data structures, which we must reuse for the reasons stated above,
    it is possible to align the Ranges data structures with the tabular notions of the dplyr API
    so we will extend dplyr to incorporate genomic notions

We have created a DSL called `plyranges` that uses the existing
`IRanges` and `GenomicRanges` packages in Bioconductor as a back-end. Our API 
still has the speed and efficiency of these packages but
with a consistent interface. The API also extends the grammar elements in `dplyr` 
for performing genomic specific manipulations such as finding overlapping 
regions or nearest neighbour regions between genomic intervals. 
The _plyranges_ DSL is specifically designed to enable fast interactive 
analysis of _Ranges_ objects but can also be used for scripting 
genomic data workflows.

@@

>>>>

There is a large variety of genomics data available to the scientific community,
ranging from highly specific sequencing assays to publicly availble collections.
These datasets similarly contain a wide variety of features: annotations of 
genomic features such as genes or exons for reference genomes, large coverage 
vectors, peaks called from ChIP-Seq experiments and many more. A modern biologist
is expected to be able to reason about, combine and wrangle these heterogenous
data sources in order to answer questions about their experimental hypotheses.
Abstractions in the forms of data representation and software interfaces are
key factors in inhibiting or enabling a biologist to construct reproducible data workflows
to answer their research questions.

@@

The Bioconductor/R packages `IRanges` and `GenomicRanges` 
define a DSL for analysing genomics data with R via
an object-oriented (OO) interface @Lawrence2013-wg; @Huber2015-ei. This interface
provides abstractions for representing genomics data and enables interoperability
by allowing users and developers to use these abstractions in their own code
and packages. However, due to the large number of abstractions there are costs
involved. Due to the OO interface each abstraction will behave slightly 
differently for any given generic method. There is a cognitive load on the user
to understand the appropriate context for each abstraction. An alternative approach
is to recognise that most genomic data analysis tasks can be performed with 
data structures that naturally represent genomic intervals. That is, genomics
data can be represented as sets of integer pairs, where each pair has a start
and an end coordianate, along with strings representing strand and chromosome 
location. Measurements or metadata can be included as additional annotations
on each pair.  As an example. a gene can be represented with its coordinates, along with its 
identifier and the identifiers of its exons; or an RNA-seq experiment may be
represented as sets of genes with a matching count column.

@@

The tidy data abstraction enables the construction of fluent application programming interfaces (APIs) [@Fowler2010-zd] that are optimised for readability. An example of a 
fluent interface built on the tidy data pattern is the R pacakge `dplyr` [@Wickham2017-dplyr]. 
The package defines a consistent syntax and style for transforming tabular data sets via
functions that are named as verbs. Each function performs an action on the data
set and is cohesive and endomorphic. Cohesion means that each function performs
a singular task and has no side-effects, while endomorphism means a function
call returns the same type of object as its input. Taken together these
features enable a user to write human readable analysis workflows.

We have created a DSL called `plyranges` that is directly motivated by the design
of `dplyr` but extends its  grammar to incorporate features of genomics data.
Our interface uses the existing Bioconductor packages `IRanges` and 
`GenomicRanges` but simplifies them by only exposing tidy genomic interval
data structures to the user. Hence, we can encapsulate the features of the 
of the `dplyr` API while performing genomic specific transformations 
such as finding overlapping  regions or nearest neighbour regions between 
genomic intervals.  The `plyranges` DSL is designed to enable fast interactive 
analysis of genomics but also enables the creation of reproducible genomic 
data workflows.

@@

<--- need to formally introduce the Ranges classes now --->
